---
title: "Iscience-PLSC-Ap-GM-Rscript"
author: "Kirthana"
date: "2023-01-17"
output:
  html_document:
    fontsize : 9 pt
    theme: united
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Import Libraries
```{r}

rm(list = ls())
graphics.off()
library(tidyverse)
library(ExPosition)
#install.packages('TInPosition') # if needed
library(TExPosition)
library(TInPosition)
library(PTCA4CATA)
#devtools::install_github('HerveAbdi/data4PCCAR')
library(data4PCCAR)
library(plyr);library(dplyr)
library(corrplot)
library(ggplot2)
library(cowplot)
library(readr)
library(gridExtra)
library(grid)
library(readxl)
library(here)
library(psych)


```


**Data :**  cleaned data, pre processed and grouped into two different datasets - one for RSI measures and other for Air pollutants exposure measures from ABCD data.
N = 8796

```{r}
# load cleaned and prepped  RSI data

# data_RSI <- RSI data

#load cleaned and prepped  Air pollutants data

# df_air_new <- Air pollutants data


```

## Performing residualization 

###  MLRM for RSI data

```{r}


# dependent variables
Group1.Y <- as.matrix(data_RSI)
dim(Group1.Y)

# independent variables/covariates
Group1.X <- df_covariates
dim(Group1.X)

# note:
# 1) check all variables- continous/factor
# 2) run str()
# 3) check for normality, homoscedascity,linearity,collinearity

# Multiple regression model
lm.Group1<- lm(Group1.Y~ Group1.X$interview_age + 
                           (Group1.X$race_ethnicity) + 
                          Group1.X$overall.income +
                          (Group1.X$demo_prnt_empl_v2)+ 
                          (Group1.X$high.educ)+
                           Group1.X$neighbor_avg_p+
                           Group1.X$sex + 
                           Group1.X$handedness +
                           Group1.X$mri_info_manufacturer + as.factor(Group1.X$site_names)+
                          Group1.X$dmri_rsi_meanmotion, na.action = na.omit) # data =)



# R.square -> how well the model explains the variation in the data which is not random
# Theorotical model performace is defined as R square

Group1_residuals <- data.frame()
Group1_residuals<- as.data.frame(lm.Group1$residuals) 
                                 



 
```

#### Measure goodness of fit

```{r echo=TRUE, fig.height=10, fig.width=12}
# for one outcome variable
# plot(lm.Group1)

# for matrix of outcome variables
# loop the variables to obtain qq plot

for (i in 1: ncol(Group1_residuals)) {
  qqnorm(Group1_residuals[,i],  main=paste0("Q-Q plot for ",colnames(Group1_residuals)[i],sep=""))
  qqline(Group1_residuals[,i])
}

hist.data.frame(as.data.frame(Group1_residuals))
# evaluation by predicton

f1 <- fitted(lm.Group1)
r1 <- residuals(lm.Group1)


for(i in 1:ncol(f1)) {
    plot(f1[,i],r1[,i] , main=paste0("fitted VS residual:   ",colnames(f1)[i],sep=""))
}


```


### MLRM for Air pollutants exposure data

```{r echo=TRUE}

#names(df_air_new)[8:10] <-  c("PM2.5", "NO2", "O3")
Group2.Y <- as.matrix(df_air_new[,8:10])
corP_pollutants <- cor(df_air_new[,8:10])
corK_pollutants<- cor(df_air_new[,8:10], method = "kendall")
corSP_pollutants<- cor(df_air_new[,8:10], method = "spearman")

lm.Group2<- lm(Group2.Y~ Group1.X$interview_age + 
                           Group1.X$race_ethnicity+
                          Group1.X$overall.income +
                          (Group1.X$demo_prnt_empl_v2)+
                          (Group1.X$high.educ)+
                           Group1.X$neighbor_avg_p+
                           Group1.X$sex +
                           Group1.X$handedness +
                           Group1.X$mri_info_manufacturer + as.factor(Group1.X$site_names)+
                          Group1.X$dmri_rsi_meanmotion,  na.action = na.omit)

Group2_residuals<- data.frame()
Group2_residuals<-as.data.frame(lm.Group2$residuals) 


```



#### Measure goodness of fit

```{r echo=TRUE}

for (i in 1: ncol(Group2_residuals)) {
  
  qqnorm(Group2_residuals[,i],  main=paste0("Q-Q plot for ",colnames(Group2_residuals)[i],sep=""))
  qqline(Group2_residuals[,i])
}

hist.data.frame(as.data.frame(Group2_residuals))

plot(fitted(lm.Group2), residuals(lm.Group2))
f <- fitted(lm.Group2)
r <- residuals(lm.Group2)

# add linear smoother 
for(i in 1:ncol(f)) {
    plot(f[,i],r[,i] , main=paste0("fitted VS residual:   ",colnames(f)[i],sep=""))
}

```

## Setting up colors for Analysis

```{r}

# setting colors for RSI brain regions
colors_brain <- c("pink","blue","yellow","deeppink","plum","skyblue","violet","darkgreen")
col4col_n0<- as.matrix(c(colors_brain,colors_brain,colors_brain,colors_brain,"orange","orange"))

col4rsi<- col4col_n0


# setting colors for air pollutants
col4air <- c("#5f9ea0","#305ABF", "#305ABF")

col4column_rsi<- matrix(nrow = ncol(data_RSI), ncol = 1)
color.vec <- as.matrix(unique(colnames(data_RSI)))
for(i in 1:length(unique(colnames(data_RSI)))){
  col4column_rsi[which(color.vec == (unique(colnames(data_RSI)))[i])] <-col4rsi[i]
}




# colors set by site
library(RColorBrewer)
color_site <- brewer.pal(4,"PuOr")
color_site<- colorRampPalette(color_site)(21)
names(color_site) <- levels(unique(df_covariates$site))
colScale <- scale_colour_manual(name = "site",values = color_site)




# Color for rows
col4row  <- matrix(nrow = nrow(df_covariates), ncol = 1)
  
for (i in 1:length(unique(df_covariates$site))) {
    col4row[which(df_covariates$site == unique(df_covariates$site)[i])] <- color_site[i]
  }


# Colors set by area
color_area <- brewer.pal(4,"PuOr")
color_area<- colorRampPalette(color_area)(3)
names(color_site) <- levels(unique(df_covariates$reshist_addr1_urban_area))
colScale <- scale_colour_manual(name = "area",values = color_site)

col4row_area  <- matrix(nrow = nrow(df_covariates), ncol = 1)
  
for (i in 1:length(unique(df_covariates$reshist_addr1_urban_area))) {
    col4row_area[which(df_covariates$reshist_addr1_urban_area == unique(df_covariates$reshist_addr1_urban_area)[i])] <- color_area[i]
  }


col4row_sex <- df_covariates$sex
col4row_sex <- as.matrix(recode(col4row_sex,
                                "F" =  "red",
                                "M" =  "green",
))

# colors set by age
df_covariates$age<-cut(df_covariates$interview_age, breaks=c(0, 114, 120, 126,133), labels=c("9-9.5 yrs", "9.5-10 yrs", "10-10.5 yrs", "10.5-11 yrs"))

col4row_age <- df_covariates$age
col4row_age <- as.matrix(recode(col4row_age,
                                "9-9.5 yrs" =  "red",
                                "9.5-10 yrs" =  "plum",
                                "10-10.5 yrs" =  "mediumorchid3",
                                "10.5-11 yrs" =  "royalblue"))

```



## PLSC Analysis

### 1. Correlation plots

```{r fig.height= 5, fig.width=20, fig.align='center'}

#Input data for PLSC
data1 <- as.data.frame(Group1_residuals)
data2 <-as.data.frame(Group2_residuals)


# Compute the covariance matrix

XY.cor.pearson <- cor(data2,data1)


corrplot(XY.cor.pearson, method = "color", tl.cex = 1, tl.col = col4rsi,
         addCoef.col = "black", number.digits = 3, number.cex = .6,
         cl.pos = 'b', cl.cex = .7,title = "Pearson Correlation",
         addCoefasPercent = F, mar=c(0,0,1,0),
         col = colorRampPalette(c("darkred", "white","midnightblue"))(6))


#XY.cor.kendall <- cor(data1, data2, method = "kendall")
# corrplot(XY.cor.kendall, method = "color", tl.cex = 1, tl.col = col4rsi,
#          addCoef.col = "black", number.digits = 3, number.cex = .6, 
#          cl.pos = 'b', cl.cex = .7, title = "Kendall Correlation",
#          addCoefasPercent = F, mar=c(0,0,1,0),
#          col = colorRampPalette(c("darkred", "white","midnightblue"))(6)) 
```



### 2. Package details


tepPLS(DATA1, DATA2, center1 = TRUE, scale1 = "SS1", center2 = TRUE, scale2 = "SS1", DESIGN = NULL, make_design_nominal = TRUE, graphs = TRUE, k = 0)

DATA1 : Data matrix 1 (X)

DATA2 : Data matrix 2 (Y)

center1 : a boolean, vector, or string to center DATA1. See expo.scale for details.

scale1 : a boolean, vector, or string to scale DATA1. See expo.scale for details.

center2 : a boolean, vector, or string to center DATA2. See expo.scale for details.

scale2 : a boolean, vector, or string to scale DATA2. See expo.scale for details.

DESIGN : a design matrix to indicate if rows belong to groups.

make_design_nominal	: a boolean. If TRUE (default), DESIGN is a vector that indicates groups (and will be dummy-coded). If FALSE, DESIGN is a dummy-coded matrix.

graphs : a boolean. If TRUE (default), graphs and plots are provided (via tepGraphs)

k	: number of components to return.



```{r echo=TRUE}

data.design <- df_covariates$site


# make the design into a vector
data.design.vec <- as.matrix(data.design)

rownames(data.design.vec) <- df_covariates$subjectkey


# Applying PLSC function to the prepped data
pls.res <- tepPLS(data1,data2, DESIGN = data.design,  make_design_nominal = TRUE, graphs = FALSE)

```

### 3. Scree Plot

The scree plot shows a weird pattern because the null hypothesis is that there is no correlation between the tables (Null = 0) 
Hence eigenvalues greater than zero become significant. 

The results of the permutation test gives us the eigenvalues.

```{r echo=TRUE}
# no.of eigenvalues
nL <- min(ncol(data1),ncol(data2))


# Applying permutation test to the input data for PLSC 
resPerm4PLSC <- perm4PLSC(data1, # First Data matrix 
                          data2, # Second Data matrix
                          permType = 'byColumns',
                          nIter = 10000 # How many iterations
                          )
print(resPerm4PLSC)


# obtaining the scree plot
my.scree <-PlotScree(ev = pls.res$TExPosition.Data$eigs,
          title = 'PLSC- Scree Plot',
          p.ev = resPerm4PLSC$pEigenvalues,
          plotKaiser = TRUE, 
          color4Kaiser = ggplot2::alpha('darkorchid4', .5),
          )


my.scree



# additional permutation histogram plots

zeDim = 1
eigs <- resPerm4PLSC$permEigenvalues[,zeDim]
obs <- as.numeric(resPerm4PLSC$fixedEigenvalues[zeDim])
pH1 <- prettyHist(
distribution = eigs, 
observed = obs,
xlim = c(0.002,0.018), # needs to be set by hand
breaks = 30,
border = "white",
main = paste0("Permutation Test "),
xlab = paste0("Inertia of samples
Latent dimension ",zeDim),
ylab = "Number of samples",
counts = F)

```

### 4. Looking at the First Pair of Latent variables by site 


```{r}
# ploting the first latent variable of data1(X) and first latent variables of data2(Y). We are tryingto see if these two latent variables are similar or not.

# first pair of latent variables:

latvar.1 <- cbind(pls.res$TExPosition.Data$lx[,1],
                  pls.res$TExPosition.Data$ly[,1])
colnames(latvar.1) <- c("Lx 1", "Ly 1")

# compute means
lv.1.group <- getMeans(latvar.1,data.design.vec)

col4Means <- as.matrix(color_site)
rownames(col4Means) <- rownames(lv.1.group)

# compute bootstrap - for confidence intervals
lv.1.group.boot <- Boot4Mean(latvar.1, data.design.vec)
colnames(lv.1.group.boot$BootCube) <- c("Lx 1", "Ly 1")

rownames(latvar.1) <- df_covariates$subjectkey

# plotiing the factor Maps
plot.lv1 <- createFactorMap(latvar.1,
                         col.points = col4row,
                         col.labels = col4row,
                         alpha.points = 0.7,
                         force = 0.01
                      
                         )

plot1.mean <- createFactorMap(lv.1.group,
                              col.points = col4Means,
                              col.labels = col4Means,
                              cex = 4,
                              pch = 17,
                              force =0.1,
                              alpha.points = 0.8)

plot1.meanCI <- MakeCIEllipses(lv.1.group.boot$BootCube[,c(1:2),], # get the first two components
                              col = col4Means[rownames(lv.1.group.boot$BootCube)],
                              names.of.factors = c("Lx 1", "Ly 1")
                              )

plot1 <- plot.lv1$zeMap_background + plot.lv1$zeMap_dots+ plot.lv1$zeMap_text + plot1.mean$zeMap_dots + plot1.mean$zeMap_text + plot1.meanCI 
plot1

# check for the outlier
outlier <- subset(latvar.1, rowSums(latvar.1 > 0.3) > 0)

indx <- which(rownames(latvar.1)== rownames(outlier))
print(rownames(outlier))

```

### 5. Obtaining Column factor scores

```{r echo=TRUE}
### Column Factor scores of the 1st component of data1 representing RSI measures and data2 representing air pollutants

#Fi:column factor scores for data1(RSI measures) or Loadings of data1(RSI measures)
Fi<- pls.res$TExPosition.Data$fi

#Fi:column factor scores for data2(Air pollutants exposure) or Loadings of data2(air pollutants exposure)
Fj<- pls.res$TExPosition.Data$fj

# Generating loadings map of Fi
p.loadings <- createFactorMap(Fi,
                              axis1 = 1,
                              axis2 = 2,
                              display.points = TRUE,
                              display.labels = T,
                              col.points = col4column_rsi,
                              col.labels = col4column_rsi,
                              title = "Loadings of Columns RSI",
                              pch = 20,
                              cex = 3,
                              text.cex = 3,
                              )

label4map <- createxyLabels.gen(x_axis = 1, y_axis = 2,
                             lambda = pls.res$TExPosition.Data$eigs,
                             tau = pls.res$TExPosition.Data$t
                             )
                             
p.plot <- p.loadings$zeMap + label4map 


p.plot

# Generating loadings map of Fj
q.loadings <- createFactorMap(Fj,
                              axis1 = 1,
                              axis2 = 2,
                              display.points = TRUE,
                              display.labels = TRUE,
                              col.points = col4air,
                              col.labels = col4air,
                              title = "Loadings of Columns Air",
                              pch = 20,
                              cex = 3,
                              text.cex = 4,
                              )

label4map <- createxyLabels.gen(x_axis = 1, y_axis = 2,
                             lambda = pls.res$TExPosition.Data$eigs,
                             tau = pls.res$TExPosition.Data$t
                             )
                             
q.plot <- q.loadings$zeMap + label4map 

q.plot


```

### 5. Column Loadings  

These are the loadings obtained after performing SVD(R) on correlation matrix.


```{r echo=TRUE}

#generating bar plots for loadings of RSI data table
P.data1<- pls.res$TExPosition.Data$pdq$p


plot_P.data1 <- PrettyBarPlot2(
                       bootratio = P.data1[,1], 
                       threshold = 0, 
                       ylim = NULL, 
                       color4bar = gplots::col2hex(col4rsi),
                       color4ns = "gray75", 
                       plotnames = TRUE, 
                       main = 'Loadings of RSI variables', 
                       ylab = " P Loadings ",
                       horizontal = TRUE)


plot_P.data1 

#generating bar plots for loadings of Air pollutants data table
Q.data2<- pls.res$TExPosition.Data$pdq$q


plot_Q.data2 <- PrettyBarPlot2(
                       bootratio = Q.data2[,1], 
                       threshold = 0, 
                       ylim = NULL, 
                       color4bar = gplots::col2hex(col4air),
                       color4ns = "white", 
                       plotnames = TRUE, 
                       main = 'Loadings of Air pollution variables', 
                       ylab = " Q Loadings ")



plot_Q.data2

# saving the loading table into excel file.
df_loadings <- as.data.frame(c(P.data1[,1],Q.data2[,1]))

names(df_loadings)<- "Loadings(Dim1)"
df_loadings <-cbind(" "=rownames(df_loadings), df_loadings)
library("writexl")
write_xlsx(df_loadings,"D:\\USC\\Project1_4.0\\Analysis\\loadings_dim1.xlsx")
```

### 6. Inference Bootstrap


The Bootstrap ratio barplot show that the contributions are significantly stable.
```{r echo=TRUE}
#Looking into what the resBootPLSC is giving us 
resBoot4PLSC <- Boot4PLSC(data1, # First Data matrix 
                          data2, # Second Data matrix
                          nIter = 10000, # How many iterations
                      Fi = pls.res$TExPosition.Data$fi,
                      Fj = pls.res$TExPosition.Data$fj,
                      nf2keep = 2.5,
                      critical.value = 2.5,
                      # To be implemented later
                      # has no effect currently
                      alphaLevel = 1)

resBoot4PLSC
```



```{r echo=TRUE}
BR.I <- resBoot4PLSC$bootRatios.i
BR.J <- resBoot4PLSC$bootRatios.j

# saving the bootrap rations into excel

df_bootstrap <-  as.data.frame(c(BR.I[,1],BR.J[,1]))

names(df_bootstrap)<- "Bootrap Ratios(Dim1)"
df_bootstrap <-cbind(" "=rownames(df_bootstrap), df_bootstrap)


write_xlsx(df_bootstrap,"D:\\USC\\Project1_4.0\\Analysis\\bootstrap_dim1.xlsx")




#bootstrap ratios for dimension 1
laDim = 1

ba001.BR1.I <- PrettyBarPlot2(BR.I[,laDim],
                        threshold = 2.5,
                        font.size = 4,
                        color4ns = "gray85", 
                   color4bar = col4rsi, # we need hex code
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle(paste0('Latent dimension ', laDim), subtitle = 'RSI Grey Matter Saliences (Loadings)')

ba002.BR1.J <- PrettyBarPlot2(BR.J[,laDim],
                        threshold = 2.5,
                        font.size = 4,
                        color4ns = "gray85", 
                   color4bar = gplots::col2hex(col4air),
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle(paste0('Latent dimension ', laDim), subtitle = 'Air pollution Saliences (loadings)')

ba001.BR1.I
ba002.BR1.J 


```

```{r echo=TRUE}
BR.I <- resBoot4PLSC$bootRatios.i
BR.J <- resBoot4PLSC$bootRatios.j


#bootstrap ratios for dimension 1
laDim = 1

ba001.BR1.I <- PrettyBarPlot2(BR.I[,laDim],
                        threshold = 3,
                        font.size = 4,
                   color4bar = col4rsi, # we need hex code
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle(paste0('Component ', laDim), subtitle = 'Table 1')

ba002.BR1.J <- PrettyBarPlot2(BR.J[,laDim],
                        threshold = 3,
                        font.size = 4,
                   color4bar = gplots::col2hex(col4air),
                  ylab = 'Bootstrap ratios'
                  #ylim = c(1.2*min(BR[,laDim]), 1.2*max(BR[,laDim]))
) + ggtitle(paste0('Component ', laDim), subtitle = 'Table 2')

ba001.BR1.I
ba002.BR1.J 

```
